% =============================================================================
% CHAPTER 2: Background
% =============================================================================
% SPDX-FileCopyrightText: 2026 Frank Langbein <frank@langbein.org>
% SPDX-License-Identifier: LPPL-1.3c
%
% Suggested sections: Context (or Wider context / Domain and context); Related
% work (or Literature review / Prior work); Problem statement (and research
% question(s)). Optional: Theory and concepts; Stakeholders and constraints
% (as subsections of Context or separate). End with a clear problem statement.
% Assume a competent reader; avoid lengthy descriptions of standard tools.
% =============================================================================

\section{Context}\label{sec:context}

Cyber security education has steadily moved towards practical scenario based learning, honing skills rather than declaring them through guided courses. Students need to be able to understand how specific vulnerabilities are exploited and work 'in the wild' rather than just understanding definitions such as SQL injection or privilege escalation. \gls{ctf} exercises have become a widely used educational tool, with prior work arguing that they complement traditional teaching and cover knowledge and skills in a way that map well to cyber security training (Švábenský et al., 2021)\cite{svabensky_automated_2024}. Within this, VulnForge attempts to address a specific aspect of this problem, whether generative AI can be used to create and manage personalised vulnerable training environments that remain educationally useful, address shortcomings of existing applications, operationally safe and practically deployable.

Existing cyber security training is dominated by static, pre-authored challenges that will often expose the learner to the same vulnerability, scenario, and flag each time. There is no mainstream platform that currently offers generative AI to synthesis vulnerabilities paired with a context aware tutor. Providing such a platform can help turn abstract concepts into simulated practice. Commercial and community platforms were built with a shared architectural assumption; once a challenge is authored, reviewed, and published, every user will encounter an identical artifact. Although this assumption provides positive operational economics, it also underlies the most commonly cited pedagogical weakness. The static challenges tend to overemphasize technical exploitation, without the user being able to explore limitation of their understanding. Take a web vulnerability, some of the web platforms may tell the user to craft a command to exploit a path traversal\cite{noauthor_cwe_nodate}, however, the user hasn't been able to yet use their knowledge of security to realise that is what they needed. 

The demand for these cyber security skills has grown considerably faster than education around them. The 2024 ISC2 Cybersecurity Workforce study estimated that the global workforce gap in relation to professionals needed to defend organisations reached 4.7 million in 2024, a 19.1 percent increase from 2023\cite{noauthor_results_nodate}. The study also reported that 30 percent of respondents reported skills gaps within their teams, and 33 percent reported staffing shortages. In addition, the United States Bureau of Labor Statistics projects a 29 percent growth rate for information security analysts through to 2034\cite{noauthor_information_nodate}, demonstrating that this is not merely a hiring problem, there is a demonstrable skill gap. 

\section{Related work}\label{sec:related}

Existing platforms have already proven the value of hands on training. Hack The Box Academy has a traditional CTF route. It presents cybersecurity training in a highly interactive process, with target hosts, checkpoints, exercises and skill assessments that allow learners to reproduce techniques they have learnt about rather than simply read about them. Katsantonis et al. 2023 \cite{katsantonis_cyber_2023} argues that cyber ranges have become important learning and training tools, but notes that their design is often hindered by a lack of shared standards and methodologies. VulnForge is not just attempting to generate vulnerable code, but rather it has a broader system, having an LLM produce structured vulnerability specifications, an orchestration layer to deploy these in isolated challenge environments, allowing the learner to access them remotely and providing a tutor to give contextual hints. The proposed solution is closer to a cyber range than a regular CTF. Lazarov et al. (2025)\cite{lazarov_lessons_2025} reinforces this idea by describing cyber ranges as controlled environments for deploying virtual machines and networks, and by having increased interactivity (the contextual tutor in this case), it can improve engagement and the development of practical skills. Therefore, the educational requirement is not only creating exploitable targets, but also environments in a system that supports learning. 

TryHackMe\cite{tryhackme} (fix reference) founded in 2018, takes a beginner oriented approach, organising content into rooms, combining guided labs with written instructions with deployable targets. Similar to Hack The Box, each deployed instance contains identical vulnerabilities and accepts the same flag, with exceptions for King of the Hill modes that rotate flags hourly whilst leaving the vulnerability unchanged. picoCTF\cite{picoctf} (fix reference), the largest free student-oriented CTF also follows a similar model, cloning vulnerable web applications for each learner, with a lack of content variation. 

Including aspects of a CTF is particularly relevant here as it provides a manageable pedagogical unit, bounded with a clear task and goal, with observable progress and explicit proof of successful exploitation. Švábenský et al. (2021)\cite{svabensky_automated_2024} describes CTFs as a popular form of modern cyber security education, having flags as a convenient proof of exploitation, with challenge categories that are mapped to a specific class demonstrates the appeal of this model. Traditional CTF content does raise questions about replay-ability, variety and personalisation, so once the users have exhausted a fixed content library, the educational value rapidly diminishes. 

\section{Procedural Challenge Generation}\label{sec:procedural}
There has been some work where existing systems have attempted to anticipate this weakness, and have some variety to CTF content. The foundational contribution here is the Automatic Problem Generation (APG) by Burkey et al(2015) at the UNSENIX Summit on Gaming, Games, and Gamificiation in Security Education\cite{burket_automatic_2015}. APG was motivated by flag sharing behavior obsereved during picoCTF in 2014. This system generates many instances of a challenge from a single parametereised template, to ensure that each player receives a unique flag, and in some tasks, a unique problem structure. However, this was not extensible to challenge types that could not be captured by a declarative template. AutoCTF took a different approach, extending the LAVA binary analysis framework to inject memory corruption bugs into existing codebases. A four university trial demonstrated that students could solve these automatically, but the technique was tighly coupled with LAVA's data flow, and did not extend beyond memory corruption bugs in C programs. 

The closest pre-cursor to wide adoption of LLMs was the Security Scenario Generator\cite{schreuders_security_nodate}(SecGen), which was a Ruby based framework that composes randomly selected Puppet modules, each encoding a specific service misconfiguration, known CVE or user account vulnerability into a virtual machine. This did support a rich variety of scenarios, and included a hints system, however, each vulnerability had to be hand-authored, with the system randomising the selection of the Puppet modules, not the vulnerability itself. APACA\cite{eckroth_alpaca_2019} extended the idea of a composable environment using AI planning over a structured vulnerability and configuration to construct multi-chain exploits, however, similar to SecGen, it relied on a pre-populated library. 

The most recent example of an LLM driven agent is AgentCyTE\cite{rodriguez_agentcyte_2025}. This uses XML files for the CORE network emulator, integrating CVEs from Vulnhub into training scenarios. This is the closest current example to the LLM generated approach that VulnForge will use. However, critically, AgentCyTE's LLM modifies scenario creation, using a library of pre-selected CVE-backed docker images; it does not synthesize new vulnerabilities. 

\section{LLMs as authors of insecure code}\label{sec:LLMs}
The empirical evidence that LLMs reliably produce security relevant code is extensive enough that it constitutes an established finding rather than a speculative one. The foundational result is from Pearce et al (2022)\cite{pearce_asleep_2021} whose systematic evaluation of Github Copilot's code contributions across 89 scenarios drawn from MITREs 2021 Top 25 Common Weakness Enumeration found that approximately 40 percent of generated programs contained exploitable vulnerabilities. This was concentrated in CWEs for buffer overlows, SQL injections and use after free. This demonstrates that LLMS, will, without special prompting insert vulnerability classes that are valuable for training purposes.  

A more concerning dimension of this behaviour was established by Perry et al (2023) \cite{perry_users_2023}, whose experiment found that 47 participants found that AI assisted programmes produced significantly less secure code, and simultaneously believed that their code was secure. This demonstrates the true implication that LLMs not only produce insecure code, but can skew developer confidence. This matters as it suggests that an LLM based training platform must not simply produce code and assume correctness, it should treat the generated output as untrusted until validated. In VulnForge, we will use a structured vulnerability schema, sever-side flag generation, deployment checks and evaluation of whether the generated challenge is actually exploitable. 

Offensive agent literature demonstrates that LLMS can not only produce vulnerable code, but also reason about it and exploit it at levels approaching human capability. PentestGPT\cite{deng_pentestgpt_2024} achieved a 228.6 percent improvement over the baseline GPT-3.5 model on penetration testing benchmarks. Cybench evaluates language model agents on professional CTF tasks and shows that frontier models can solve some easier tasks, especially when providing with scaffolding or subtasks (Zhang et al., 2025)\cite{zhang_cybench_2025}. NYU CTF Bench similarly evaluates models against CTF like challenges, and had similar findings(Shao et al., 2024)\cite{shao_nyu_2025}, but did find that LLMs have certain limitations related to certain types of challenges such as incident response (IR). These systems are relevant as VulnForge will have the model playing two roles, the author and the tutor. The model creates the vulnerability, but it also must retain enough context to give hints. This helped influence the design  decisions, the LLM should not output arbitrary prose, but instead a schema constrained vulnerability containing specifics relating to services, files, packages and finally produce a solution summary. 

\section{LLM tutoring}\label{sec:llm-tutoring}
A tutor is used as part of this project as practical cyber security tasks can quickly become frustrating for a learner when they are stuck without feedback. A static hint list can reveal too much, and a full walk-through undermines the learning objective. An adaptive tutor can provide progressive scaffolding, first understanding what the user has tried, then suggesting different enumeration strategies, and explaining relevant concepts when needed. 

The idea of a tutor to aid learning is supported in educational psychology. Bloom et al, (1984) \cite{bloom_2_1984} argued that one-to-one tutoring can produce more significant learning gains in comparison with conventional classroom instructions. Students receiving tutoring achieved two standard deviations higher than those in conventional classes, with 90 percent of tutored students achieving mastery that only 20 percent of conventional students achieved [pp. 1-12].  VahLehn et al, (2011)\cite{vanlehn_relative_2011} reviewed intelligent tutoring systems and found that well-designed tutoring systems can produce the effectiveness of human tutoring in some cases. Although these results may not directly transfer to CTFs, they support the general educational principle; timely, adaptive feedback can be useful to learners attempting difficult tasks. 

Lewis et al, (2020) \cite{lewis_retrieval-augmented_2021} claims that Retrieval Augmented Generation (RAG) is commonly used to ground responses in context rather than relying on model memory. VulnForge will use a similar idea, the tutor will be  grounded in the generated challenge metadata, including the attack vector, hint and solution summary. This allows it to provide challenge specific guidance rather than just generic advice about vulnerabilities. 

A risk with the tutor is revealing too much. Liffton et al, (2023)\cite{liffiton_codehelp_2024} wrote about CodeHelp, a guard-railed LLM for programming education, that gives guidance without providing the solution. VulnForge aims to take a similar approach, it should scaffold reasoning without giving the flag or exact exploit sequence. 

Prompt injection is another concern with having a chatbot. OWASP lists prompt injection in the OWASP TOP 10 for LLM Applications\cite{editor_owasp_nodate}. Wei et al (2023)\cite{wei_jailbroken_2023} shows that models can fail safety training when instructions conflict or when generalisations do not match the intended policy. In this type of platform, users are adversaries by design, they are expected to bypass restrictions and test assumptions, so there needs to be strong grounding in the system prompt and guardrails to prevent this. The implementation will need to combine prompt level rules, output filtering, application side checks and more. The tutor should be instructed not to share the complete solution, and there should be application logic to check for forbidden flag patterns. This does not eliminate prompt injection risk, but does allow the application to consider the security requirement. 

\section{Sandboxed deployment}\label{sec:sandboxed}
Running environments that are deliberately insecure present a host of security problems. The environment has to be vulnerable enough where the learner is able to exploit it, but not so exposed that a successful vulnerability is able to escape the intended boundary and attack other systems on the network or host. This makes the development of VulnForge different to simply a web application, it has to make many considerations for creating targets that enable this adversarial action. 

NIST SP 800-190 identifies container images, registries, orchestrators, runtimes and hosts as distinct areas of container security concern (Souppaya et al, 2017]\cite{}. This presents a lesson for the development of this application, the containers should not be automatically treated as having strong security boundaries. Sultan et al, (2019)\cite{} similarly described container security as a layered problem, involving kernel configuration, namespaces, mandatory access control and runtime configuration. 

This matters as the problem intends to use Proxmox VE to host the unprivileged containers. It will use Ansible for container lifecycle and vulnerability injection. LXC containers are appropriate for the scale of the project as they are lightweight, scriptable and integrate well with Proxmox VE, however they do not have the same boundaries as a regular virtual machine (VM). Standard VMs do not have a shared kernel or namespace, providing additional isolation to attacks. A production ready version of this application would require stronger isolation, more network policies, aggressive egress control, and possibly technologies such as Firecracker\cite{noauthor_firecracker_nodate} which is purpose built for managing secure multi-tenant containers.

VulnForge does not attempt to solve mutli-tenant sandboxing at a commercial scale, instead it aims to offer some safety requirements, with carefully chosen architecture to provide some network and host isolation, without significantly impacting developmental cost or time. 

Ephemerality is something that will be carefully considered as part of the development. If a container remains online continually, it becomes a standing target, taking up extra resources, and opening up the host to attack. Time limited environments that clean up resources reduce this risk and also act as a resource control for users, preventing them from spinning up too many containers causing excessive load on the service. Network isolation is also considered here, the containers should not be used as a pivot point to the wider internet, not be used to accessed other learners containers. 


\section{Problem statement}\label{sec:problem}

The background discussion demonstrates the clear problems within this domain, it has proven that cyber security labs and AI tutors give real value to the end user. But the problem lies in whether they can be combined into a coherent, safe, and educationally useful system that generates vulnerable, isolated environments on demand and then exposes them through controlled infrastructure. From here, it should support the learner with contextual guidance and reliably clean-up the environment afterwards. Put simply, VulnForge will be investigating whether generative AI can move from just being an accessory that people use, to being part of the content production and delivery pipeline itself. 

This leads to the following questions:
\begin{enumerate}
    \item To what extent can a \gls{llm} pipeline generate vulnerability specifications that are both deployable and sufficiently diverse to support cyber security training within defined challenge categories?
    \item How can generated challenges be deployed and exposed through isolated, short-lived environments without undermining host safety , challenge integrity or resource control?
    \item Relative to a purely curated challenge library, does an on-demand generated approach offer meaningful improvement in re-playability and education?
\end{enumerate}