% =============================================================================
% CHAPTER 3: Methodology (or ``Specification, Design and Implementation'' / ``Approach'')
% =============================================================================
% SPDX-FileCopyrightText: 2026 Frank Langbein <frank@langbein.org>
% SPDX-License-Identifier: LPPL-1.3c
%
% This chapter describes how your solution works. Title it Methodology, Approach,
% or Specification, Design and Implementation; adapt or split as needed. Not all
% sections below apply to every project---pick, omit, or rename.
%
% Suggested sections by project type:
%   System design:  Specification, System design, Implementation. Optional: Datasets
%                   if data is central; Ethical considerations.
%   Research:       Research design, Literature and evidence (search strategy,
%                   inclusion criteria), Specification (= problem + requirements),
%                   Approach and solution, Implementation or experimental setup.
%   AI/ML:          Specification, Datasets, AI/ML models, System design (if a
%                   larger system), Implementation. Optional: Research design,
%                   Literature and evidence; Summary.
% =============================================================================

This chapter describes how the problem was addressed. Choose sections that fit your project type; omit or merge as needed. Replace with enough detail for reproduction.


\section{Research design}\label{sec:research-design}
% Research: overall approach, methods (e.g.\@ quantitative, qualitative, design science), rationale. Omit for pure system-design projects.


\section{Specification}\label{sec:spec}
% System design: requirements (what the system must do; optional subsections: functional, non-functional). Research: problem and what is required of a solution.

VulnForge has to solve three key, connected problems. Firstly, it must provide a usable web interface that allows users to create an account, store an SSH public key, choose a vulnerability category and launch a challenge. Secondly, it must safely orchestrate infrastructure code: create an isolated container, inject the user's SSH key, apply an LLM generated vulnerability specification, and then clean up after usage. Thirdly, it must integrate a generative AI tutor in a constrained way so that the LLM produces useful challenge context, without exposing the answer directly. 

\subsection{Scope and assumptions}\label{sec:scope}

The system is designed as a controlled environment for educational use, not for public scale internet usage. A user interacts with a web dashboard with their challenge machine through SSH (or a web-interface when appropriate for the challenge). The vulnerable environment will be a Ubuntu 24.04 LXC container hosted on a Proxmox VE hypervisor. The container is intended to be ephemeral, to be destroyed either after a time expiration, the challenge is completed or if it is manually deleted.  

The main operating assumptions are the following:
\begin{itemize}
    \item Users authenticate through the web application before starting a challenge
    \item Each user stores a SSH public key in their account settings before starting a challenge
    \item Students connect to a challnege as a non-root Linux user called `student`
    \item The `student` user has no default sudo access, any privilege escalation must be gained through the generated vulnerability
    \item Each user must have at most one challenge active at any time
    \item The platform stores the flag server side, and does not send it to the user
    \item The vulnerable container is reachable only through the intended lab path
\end{itemize}

\section{Datasets}\label{sec:datasets}
% AI/ML / empirical: data sources, preprocessing, train/validation/test split. Omit if not data-driven.


\section{AI and machine learning models}\label{sec:aiml}
% AI/ML: model choice, training and validation, evaluation methodology. Omit otherwise.


\section{System design}\label{sec:design}
% System design: architecture, data flow, behaviour (UML, ER, state charts). Use diagrams. Research: use ``Approach and solution'' for your proposed solution(s).


\section{Implementation}\label{sec:impl}

\subsection{Introduction}

The implementation of this system utilises ephermeral Linux containers, injecting intentional vulnerabilites into them, exposing them to a student through a controlled network path, verifiying the challenge through a flag, and prov

\subsection{Overall System Architecture}
\subsection{Infrastructure orchestration and ephemeral lab provisioning}
\subsubsection{Proxmox and container model}
\subsubsection{Ansible as the orchestration layer}
\subsubsection{Network access and learner connectivity}
\subsubsection{Automatic cleanup and the TTL reaper}
\subsection{LLM driven vulnerability generation and injection}
\subsection{Web application}




\section{Summary}\label{sec:summary}
% Optional: short summary of the methodology. Omit if the chapter is already concise.
